---
title: "STA3115_homework1"
author: "2021122006 Jaehun Shon"
date: "2024-09-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(readr);library(tidyverse);library(ggplot2);library(patchwork);library(gridExtra);library(knitr)
billboard <- read_csv("billboard.csv",
                      col_types = cols(time = col_character()))
billboard
```

# Problem 1: Splitting the "time" variable

```{r, message=FALSE}
billboard %>% 
  separate(time, into = c("minutes", "seconds"), sep = ":") %>% 
  .[, c("year", "artist.inverted", "track", "minutes", "seconds")] -> billboard.1

billboard.1 %>%
  head(10) %>%
  kable(col.names = c("Year", "Artist", "Track", "Minutes", "Seconds"),
        caption = "Top 10 rows of Billboard Data",
        format = "markdown")
```

# Problem 2: Creating a new time variable and histogram

```{r, message=FALSE}
billboard.1 |> 
  mutate(time_in_min = as.numeric(minutes) + as.numeric(seconds) / 60) -> timevar

t <- timevar$time_in_min
description <- data.frame(
  Statistic = c("Mean", "Sd", "Median", "IQR", "Min", "Max"),
  Value = c(mean(t), sd(t), median(t), IQR(t), range(t)) %>% round(., 3)
) %>% t

histplot <- ggplot(timevar, aes(x = time_in_min)) +
  geom_histogram(fill = "skyblue", color = "blue", binwidth = 0.35) +
  labs(x = "Length of the song in minutes",
       y = "Frequency")

table_grob <- tableGrob(description)
final_plot <- grid.arrange(histplot, table_grob, nrow = 2, heights = c(3/4, 1/4))
```
 The characteristic of distribution is stated in above. The data has right-skewed distribution, which can easily found in histogram.

# Problem 3: Summary statistics by weeks on chart

```{r, message=FALSE}
billboard %>%
  separate(time, into = c("minutes", "seconds"), sep = ":") %>%
  select(year, artist.inverted, minutes, seconds, x16th.week) %>%
  mutate(time_in_min = as.numeric(minutes) + as.numeric(seconds) / 60) -> billboard.2

a <- billboard.2 %>% .[is.na(.[, 'x16th.week']),] %>% pull(time_in_min) # below 15
b <- billboard.2 %>% .[!is.na(.[, 'x16th.week']),] %>% pull(time_in_min) # more than 15

compare.dt <- tibble(
  Statistic = c("Mean", "Standard Deviation", "Median", "IQR"),
  `Below 15 weeks` = c(mean(a), sd(a), median(a), IQR(a)),
  `More than 15 weeks` = c(mean(b), sd(b), median(b), IQR(b))
)

compare.dt %>%
  kable(col.names = c("Statistic", "Below 15 weeks", "More than 15 weeks"),
        caption = "Comparison of Statistics for Below 15 weeks and More than 15 weeks in song length",
        format = "markdown")
```

# Problem 4: Comparing song length distribution

```{r, message=FALSE}
plot1 <- ggplot(billboard.2, aes(x = time_in_min)) +
  geom_density(aes(fill = ifelse(!is.na(x16th.week), "More than 15 weeks", "Below 15 weeks"),
                   color = ifelse(!is.na(x16th.week), "More than 15 weeks", "Below 15 weeks")),
               alpha = 0.7, adjust = 0.7) +
  labs(x = "Length of the song in minutes", y = "Density", 
       fill = "", color = "") +
  scale_fill_manual(values = c("More than 15 weeks" = "skyblue",
                               "Below 15 weeks" = "pink")) +
  scale_color_manual(values = c("More than 15 weeks" = "blue",
                                "Below 15 weeks" = "red")) +
  theme_minimal()


qq.df <- as.data.frame(qqplot(a, b, plot.it = FALSE))
plot2 <-ggplot(qq.df, aes(x = x, y = y)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  labs(x = "Below 15 weeks", y = "More than 15 weeks") +
  xlim(range(c(a, b))) +
  ylim(range(c(a, b))) +
  theme_minimal()

pp <- plot1 + plot2
pp
```

## Density plot(left), QQ plot(right)
 
1. As density plot shows smooth distribution shape of the data, it is easy to compare between two groups.
2. I wanted to show the overall shape in easy-to-see way with a density plot, and show how further difference between groups occured with more precise by QQ plot.

# Problem 5: Comparing to the Normal distribution

```{r, message=FALSE}
qq_plot <- ggplot() +
  stat_qq(data = billboard.2 %>% filter(!is.na(x16th.week)),
          aes(sample = time_in_min)) +
  stat_qq_line(data = billboard.2 %>% filter(!is.na(x16th.week)),
               aes(sample = time_in_min), linewidth = 0.5) +
  geom_abline(aes(intercept = mean(b), slope = 0), color = 'red', linewidth = 0.1) +
  labs(x = "theoretical", y = "sample", title = 'Normal qq plot')
  
hist_plot <- ggplot() +
  geom_histogram(data = billboard.2 %>% filter(!is.na(x16th.week)), 
                 aes(x = time_in_min),
                 binwidth = 0.5, fill = "skyblue", color = "blue") +
  coord_flip() +
  labs(x = "", y = "time in minutes", title = 'Histogram  (More than 15 weeks)')

qq_plot + hist_plot
```

```{r, message=FALSE}
qq_plot <- ggplot() +
  stat_qq(data = billboard.2 %>% filter(is.na(x16th.week)),
          aes(sample = time_in_min)) +
  stat_qq_line(data = billboard.2 %>% filter(is.na(x16th.week)),
               aes(sample = time_in_min), linewidth = 0.5) +
  geom_abline(aes(intercept = mean(a), slope = 0), color = 'red', linewidth = 0.1) +
  labs(x = "theoretical", y = "sample", title = 'Normal qq plot')
  
hist_plot <- ggplot() +
  geom_histogram(data = billboard.2 %>% filter(!is.na(x16th.week)), 
                 aes(x = time_in_min),
                 binwidth = 0.5, fill = "pink", color = "red") +
  coord_flip() +
  labs(x = "", y = "time in minutes", title = "Histogram  (Below 15 weeks)")

qq_plot + hist_plot
```

1. Distribution of each group shows deviation from the normal distribution when the length of song exceeds approximately 4.5 minutes. The red line represents the mean of the length of song respectively, it shows the average value is deviated from the midpoint of the entire data range.
2. To show the skewedness of data in each group easily, histogram was added next to the QQplot. They look quite similar to the plot in Promblem 2, which is right-skewed.